---
title: A Study of War
author: Adam Austin
output: html_document
---

```{r setup, include = FALSE}
library(rprojroot)
library(parallel)
library(ggplot2)
library(bench)
library(tidyr)

root_dir <- find_root(has_dir(".git"))
war_file <- file.path(root_dir, "cards-war", "war.R")

source(war_file)

low_round_threshold <- 10

set.seed(20201124)
```


## Playing the game

To simulate a game, we choose the number of ranks and suits in our deck
(the default is 13 ranks and 4 suits, as in a standard 52-card deck).

The output is the number of rounds played in that game (_i.e._ the number of
times cards are collected by one player) and the overall winner of that game.

```{r playing}
simulate_game(ranks = 13, suits = 4)
```

The game can easily take an hour or more when playing in real life.  In this
simulation, it takes milliseconds.

```{r timing}
benchmark <- mark(simulate_game(), iterations = 1000)
plot(benchmark)
```

## Statistics of multiple games

We simulate many games and examine the distribution of winners and rounds per game.

```{r simulate}
games <- mclapply(1:10000, function(x) simulate_game(), mc.cores = detectCores() - 1)
games_data <- do.call(rbind, games)
head(games_data)
```


### How many rounds does it take to end the game?

```{r round-count, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(games_data, aes(x = round_count)) +
  geom_histogram() +
  labs(x = "rounds played in a game", y = "count") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor.x = element_blank()) +
  scale_x_continuous(breaks = seq(0, max(games_data$round_count), by = 100))
```


Some games are over in very few rounds.  What do players' starting hands looks like, after
distributing the cards, when the game only lasts `r low_round_threshold` rounds?

A deck is represented as a sampled ("shuffled") vector of integers standing
for the card ranks (2 through Ace are stored as 1 through 13), with repeating
ranks for the suits (e.g. there are 4 copies of rank 10).

```{r low-round-threshold, echo = FALSE}
round_count <- 9999
while(round_count > low_round_threshold) {
  deck <- shuffle_deck()
  player_hands <- deal_player_hands(deck)
  n_rounds <- 0
  
  while (at_least_one_card_in(player_hands)) {
    player_hands <- play_round(deck, player_hands)
    n_rounds <- n_rounds + 1
  }

  round_count <- n_rounds
}

player_hands <- deal_player_hands(deck)

cat(paste("This game ended in", round_count, "rounds with player",
    which.max(lengths(player_hands)), "winning.\n"))

data.frame(player_1 = deck[player_hands$p1],
           player_2 = deck[player_hands$p2])
```


### What proportion of games does each player win?

```{r winner-prop, echo = FALSE, warning = FALSE}
ggplot(games_data, aes(x = factor(winner))) +
  geom_bar(width = 0.5) +
  labs(x = "winning player", y = "number of games") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor.y = element_blank())
```


## How does the type of deck influence the game's length?

A standard deck has 13 ranks with 4 suits each.  As a useless thought exercise,
what if we vary this?  How does it affect the median game length (in number of
rounds)?  Here we simulate 200 games with each combination of ranks and suits.

(Note: the total number of cards must be even.)

```{r deck-type, echo = FALSE}
ranks <- seq(8, 18, by = 2)
suits <- 2:6

trials <- setNames(expand.grid(ranks, suits), nm = c("ranks", "suits"))
median_rounds <- integer(nrow(trials))

for(i in 1:nrow(trials)) {
  games <- mclapply(1:200,
                    function(x) simulate_game(ranks = trials$ranks[i], suits = trials$suits[i]),
                    mc.cores = detectCores() - 1)
  games_data <- do.call(rbind, games)
  median_rounds[i] <- as.integer(median(games_data$round_count))
  rm(games, games_data)
  gc()
}

trials$median_rounds <- median_rounds
trials$card_count <- trials$ranks * trials$suits
trials$rounds_per_card <- trials$median_rounds / trials$card_count

ggplot(trials, aes(x = suits, y = ranks, fill = median_rounds)) +
  geom_tile() +
  scale_fill_continuous(type = "viridis", name = "median\nrounds") +
  labs(x = "number of suits", y = "number of ranks") +
  scale_y_continuous(breaks = ranks) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```


What if we normalize the round count by the number of cards in the deck?

```{r deck-type-normalized, echo = FALSE}
ggplot(trials, aes(x = suits, y = ranks, fill = rounds_per_card)) +
  geom_tile() +
  scale_fill_continuous(type = "viridis", name = "median\nrounds\nper card") +
  labs(x = "number of suits", y = "number of ranks") +
  scale_y_continuous(breaks = ranks) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```